---
title: "scGAS Workflow: Gene Activation Potential from scATAC-seq"
author: "Weiyan Chen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{scGAS Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # set to TRUE only in a live environment
)
```

## Overview

**scGAS** infers Gene Activation Potential (GAS) from single-cell ATAC-seq
(scATAC-seq) data.  The framework:

1. Maps raw scATAC-seq fragments onto a reference set of 500 bp chromatin
   accessible bins (cCREs) derived from 124 ENCODE DNase-seq samples.
2. Uses statistically significant cCRE-to-Gene associations from 167 paired
   ENCODE bulk DNase-seq / RNA-seq samples to train a per-gene **Lasso
   regression model**.
3. Aggregates cells into **Metacells** (high-resolution clusters ≥ 100 cells)
   to overcome single-cell sparsity, applies the Lasso models to obtain
   Metacell-level GAS, then propagates GAS to **single-cell resolution** via
   Gaussian-kernel initialisation and random-walk network propagation.
4. Links chromatin activation to transcriptional state through a
   **Chromatin Potential Field** that predicts differentiation trajectories.

---

## Installation

```{r install}
# Install from GitHub
remotes::install_github("ChenWeiyan/scGAS")

# Bioconductor dependencies
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("Signac", "EnsDb.Hsapiens.v75", "SCAVENGE"))
```

---

## Step 0 — Prepare inputs

### Required files

| File | Description |
|------|-------------|
| `hg19_500bp_CRE.bed` | Reference cCRE BED (provided with the package or downloadable) |
| `*_fragments.tsv.gz` | Per-replicate fragment files from Cell Ranger ATAC or multiome |
| `ENCODE_167Tissue_SigAss_pVal.05.rds` | Pre-computed cCRE-Gene association list |
| `ENCODE_167Tissue_RNA_LogNorm.rds` | ENCODE bulk RNA-seq log-normalised matrix |
| `ENCODE_167Tissue_CRE_LogNorm.rds` | ENCODE bulk DNase-seq log-normalised matrix |

The three `.rds` files are available from the scGAS project data repository
(see README for the download link).

### Gene annotations

```{r annotations}
library(EnsDb.Hsapiens.v75)
library(Signac)

annotation <- GetGRangesFromEnsDb(EnsDb.Hsapiens.v75)
seqlevelsStyle(annotation) <- "UCSC"
genome(annotation) <- "hg19"
```

---

## Step 1 — Preprocess scATAC-seq data (`scgas_preprocess`)

`scgas_preprocess()` handles fragment loading, feature-matrix construction,
QC metric computation, TF-IDF / LSI dimensionality reduction, and
high-resolution clustering.

```{r preprocess}
library(scGAS)

obj <- scgas_preprocess(
  fragment_paths     = c("rep1_atac_fragments.tsv.gz",
                         "rep2_atac_fragments.tsv.gz"),
  cre_bed_path       = "hg19_500bp_CRE.bed",
  annotation         = annotation,
  replicate_ids      = c("cd34_multiome_rep1#", "cd34_multiome_rep2#"),
  lsi_dims           = 2:30,
  cluster_resolution = 1.4,   # ≥1.4 recommended; lower if clusters < 100 cells
  min_cells_per_cluster = 100,
  n_cores            = 8,
  verbose            = TRUE
)
```

> **Tip — multiome data:** If you have matched RNA, supply `rna_matrix` and
> `meta_data` directly:
> ```{r preprocess_multiome, eval=FALSE}
> obj <- scgas_preprocess(
>   ...,
>   rna_matrix = scRNA_sparse_matrix,
>   meta_data  = cell_metadata_df
> )
> ```

Inspect the QC distribution before proceeding:

```{r qc_plot}
library(Seurat)
VlnPlot(obj, features = c("nFeature_ATAC", "TSS.enrichment",
                           "nucleosome_signal"), ncol = 3, pt.size = 0)
```

---

## Step 2 — Train Lasso models and compute Metacell GAS (`scgas_train_models`)

```{r train}
# Load pre-computed ENCODE reference data
sig_ass  <- readRDS("ENCODE_167Tissue_SigAss_pVal.05.rds")
rna_enc  <- readRDS("ENCODE_167Tissue_RNA_LogNorm.rds")
dnase_enc <- readRDS("ENCODE_167Tissue_CRE_LogNorm.rds")

models <- scgas_train_models(
  seurat_obj           = obj,
  sig_associations     = sig_ass,
  encode_rna_lognorm   = rna_enc,
  encode_dnase_lognorm = dnase_enc,
  cre_bed_path         = "hg19_500bp_CRE.bed",
  min_associations     = NULL,   # NULL → use median as threshold
  n_cores              = 8,
  verbose              = TRUE
)

# How many genes were modelled?
length(models)
```

The returned list carries the trained model for each gene plus the
normalised Metacell ATAC matrix as an attribute — passed automatically to
`scgas_compute()`.

---

## Step 3 — Propagate GAS to single-cell resolution (`scgas_compute`)

```{r compute}
scgas_mat <- scgas_compute(
  seurat_obj     = obj,
  trained_models = models,
  lsi_dims       = 2:30,
  knn_k          = 30,
  seed_fraction  = 0.10,   # top 10 % of cells used as propagation seeds
  gamma          = 0.05,
  n_cores        = 8,
  verbose        = TRUE
)

dim(scgas_mat)   # genes × cells
```

Add the scGAS matrix back to the Seurat object and run dimensionality
reduction:

```{r add_assay}
obj <- scgas_add_assay(
  seurat_obj        = obj,
  scgas_mat         = scgas_mat,
  run_dim_reduction = TRUE,
  n_pcs             = 30,
  n_neighbors       = 30
)

# Visualise key marker genes
library(Seurat)
FeaturePlot(obj, features = c("GATA1", "MS4A1", "HOXA9", "CEBPA"),
            reduction = "scgasumap", ncol = 2)
```

---

## Step 4 — Chromatin Potential Field (`scgas_chromatin_potential`)

The CPF predicts differentiation flow by linking each cell's scGAS profile to
its most transcriptomically correlated neighbours.

```{r cpf}
cpf <- scgas_chromatin_potential(
  seurat_obj   = obj,
  n_hvg        = 2000,
  knn_k        = 10,
  n_cores      = 8,
  arrow_scale  = 1.5,
  point_colour = "celltype",   # metadata column
  title        = "CD34+ BM Chromatin Potential Field"
)

# Print the vector-field plot
print(cpf$plot)

# The underlying data frame
head(cpf$cpf_df)
```

To highlight a specific lineage, subset the gene set used for CPF:

```{r cpf_lineage}
lineage3_genes <- c("HBD", "HBB", "GATA1", "KLF1")   # erythroid markers

cpf_ery <- scgas_chromatin_potential(
  seurat_obj   = obj,
  n_hvg        = 500,
  knn_k        = 10,
  title        = "Erythroid lineage — CPF"
)
print(cpf_ery$plot)
```

---

## Session information

```{r session}
sessionInfo()
```
